---
title: "MATH 448"
author: "Danyl Kuzmychov"
date: "Oct 22nd, 2021"
output: pdf_document
---


Step 1;

-- Look at data.

```{r}
library(dplyr)
library(stringr)
library(data.table)
library(ggplot2)
library(glmnet)
library(pls)
library(splines)
library(gam)
library(randomForest)

data_start = read.csv("D:/448 Project Data/Enviroment_2/GlobalLandTemperaturesByState.csv")

data_start1 = read.csv("D:/448 Project Data/Enviroment_2/GlobalTemperatures.csv")
```


Step 2;

-- Process all California data.

```{r}
ls(data_start1)

cali_temp = subset(data_start, data_start$State == "California")

global_temp_1850 = data_start1[-c(1:1188,3166:3192),]

cali_final_data = merge(cali_temp, global_temp_1850, by = "dt") # Cali Temp from 1850-2013

cali_final_data = cali_final_data[-c(1,2,3,4,5,6,7,8,9,10,11,12),]

cali_final_data$year <- substr(cali_final_data$dt,1,4)
cali_final_data$month <- substr(cali_final_data$dt,6,7)

y <- cali_final_data$AverageTemperature[4:nrow(cali_final_data)]
year <- cali_final_data$year[4:nrow(cali_final_data)]
month <- cali_final_data$month[4:nrow(cali_final_data)]
lag1 <-  cali_final_data$AverageTemperature[3:(nrow(cali_final_data)-1)]
lag2 <-  cali_final_data$AverageTemperature[2:(nrow(cali_final_data)-2)]
lag3 <-  cali_final_data$AverageTemperature[1:(nrow(cali_final_data)-3)]
aveTempSD <- cali_final_data$AverageTemperatureUncertainty[4:nrow(cali_final_data)]
aveLandTemp <- cali_final_data$LandAverageTemperature[4:nrow(cali_final_data)]
aveLandTempSD <- cali_final_data$LandAverageTemperatureUncertainty[4:nrow(cali_final_data)]

data <- data.frame(y,year,month,lag1,lag2,lag3,aveTempSD,aveLandTemp,aveLandTempSD)
data$month <- as.factor(data$month)
data$year <- as.numeric(as.character(data$year))

summary(data)
```


Goal; Try to narrow down the data by looking at the summers and winters of each year and trace their average from 1849 to 2013. 


Modeling;

-- Linear Model.

```{r}
row.number <- sample(1:nrow(data), 0.8*nrow(data))
train_lm = data[row.number,]
test_lm = data[-row.number,]

lm_1 = lm(y ~ ., data = train_lm)
lm.pred = predict(lm_1, newdata = test_lm)
summary(lm_1)
mean((lm.pred-test_lm$y)^2)

#ggplot(train_lm, aes(y)) + geom_density(fill="blue")                             #bimodal
```



Subset selection; 

-- Forward Step-wise and Backward Step-wise. 

```{r}
lm.all = lm(y ~ ., data = train_lm)                                #lm w/ all predictors
summary(lm.all)

lm.intercept = lm(y ~ 1, data = train_lm)                          #lm w/ only intercept
summary(lm.intercept)

forward_step = step(lm.intercept, direction = "forward", scope = formula(lm.all), trace = 0)     
forward_step$anova
forward_step$coefficients
lm.pred1 = predict(forward_step, newdata = test_lm)
mean((lm.pred1-test_lm$y)^2)                                                     


backward_step = step(lm.all, direction = "backward", scope = formula(lm.all), trace = 0)    
backward_step$anova
backward_step$coefficients
lm.pred2 = predict(backward_step, newdata = test_lm)
mean((lm.pred2-test_lm$y)^2)                                                        
```



Regularization;

-- Lasso Regression

```{r}
x_lasso = model.matrix(y ~ ., data = data)[,-1]        
y_lasso = data$y
dim(x_lasso)

train_data = row.number
test_data = (-train_data)
y.test = y_lasso[test_data]
grid = 10^seq(10,-2,length=100)

lasso.mod = glmnet(x_lasso[train_data,],y_lasso[train_data], alpha=1, lambda=grid)
plot(lasso.mod)                                                                        #coefficient plot

set.seed(10)

cv.out = cv.glmnet(x_lasso[train_data,],y_lasso[train_data], alpha=1)
plot(cv.out)                                                                           #CV error plot

bestlambda = cv.out$lambda.min
bestlambda                                                                            #min cross validation error

lasso.pred=predict(lasso.mod,s=bestlambda,newx=x_lasso[test_data,])
mean((lasso.pred-y.test)^2)                                                            #test MSE = 1.342475           

out = glmnet(x_lasso,y_lasso,alpha=1,lambda=grid)
lasso.coef = predict(out,type="coefficients",s=bestlambda)
lasso.coef
```



-- Ridge Regression

```{r}
x_ridge = model.matrix(y ~ ., data = data)[,-1]                                 
y_ridge = data$y
dim(x_ridge)

train_data1 = row.number
test_data1 = (-train_data1)
y.test1 = y_ridge[test_data1]

ridge.mod = glmnet(x_ridge,y_ridge,alpha=0,lambda=grid)
plot(ridge.mod)                                                                    #coefficient plot

set.seed(1)

cv.out1 = cv.glmnet(x_ridge[train_data1,],y_ridge[train_data1],alpha=0)
plot(cv.out1)                                                                      #CV error plot

bestlam1=cv.out1$lambda.min
bestlam1                                                                        #min cross validation error

ridge.pred=predict(ridge.mod,s=bestlam1,newx=x_ridge[test_data1,])
mean((ridge.pred-y.test1)^2)                                                     

out1 = glmnet(x_ridge,y_ridge,alpha=0)
predict(out1,type="coefficients",s=bestlam1)
```




Higher-Dimensional Analysis;

-- PLS Regression

```{r}
x_pls = model.matrix(y ~ ., data = data)[,-1]
y_pls = data$y

train_pls = row.number
test_pls = (-train_pls)
y.test2 = y_pls[test_pls]

set.seed(1)

pls.fit=plsr(y ~ ., data = data, subset = train_pls, scale=TRUE, validation="CV")
summary(pls.fit)                                                                        #50% of variation can be explained by 5 components
coefficients(pls.fit)

validationplot(pls.fit, val.type="MSEP")                                                #M = 9-18
validationplot(pls.fit, val.type="R2") 

pls.pred = predict(pls.fit, x_pls[test_pls,],ncomp=15)
mean((pls.pred-y.test2)^2)                                                              
```


-- PCR Regression

```{r}
x_pcr = model.matrix(y ~ ., data = data)[,-1]
y_pcr = data$y

train_pcr = row.number
test_pcr = (-train_pcr)
y.test3 = y_pcr[test_pcr]

set.seed(11)

pcr.fit = pcr(y ~ ., data=data , subset = train_pcr, scale=TRUE, validation="CV")
summary(pcr.fit)                                                                          #50% of variation can be explained by 4 components
coefficients(pcr.fit)

validationplot(pcr.fit, val.type="MSEP")                                                  #M = 15-18
validationplot(pcr.fit, val.type="R2")

pcr.pred = predict(pcr.fit,x_ridge[test_pcr,],ncomp=18)
mean((pcr.pred-y.test3)^2)                                                                #test MSE = 1.41556 Lowest at M = 18
```





Random Forests;  !! REVIEW !!

```{r}
x_forests = model.matrix(y ~ ., data = data)[,-1]
y_forests = data$y

train_forests = row.number
test_forests = (-train_forests)
y.test4 = y_forests[-train_forests]

set.seed(1)
rf.data = randomForest(y ~ ., data=data, subset = train_forests, mtry=3, importance=TRUE)
yhat.rf = predict(rf.data, newdata = data[-train_forests,])
mean((yhat.rf-y.test4)^2)

importance(rf.data)
varImpPlot(rf.data)
```




GAM;  !! REVIEW !!

```{r}
gam1=lm(y ~ month + ns(lag1,5) + ns(lag2,5) + ns(lag3,5) + ns(aveLandTemp,5), data=train_lm)
gam.pred = predict(gam1, newdata = test_lm)
mean((gam.pred-test_lm$y)^2)


summary(gam1)
```




Comparison of Methods;

```{r}
lm.mse = mean((lm.pred-test_lm$y)^2)                                            
forward.mse = mean((lm.pred1-test_lm$y)^2)                                
backward.mse = mean((lm.pred2-test_lm$y)^2)                              
ridge.mse = mean((ridge.pred-y.test1)^2)
lasso.mse = mean((lasso.pred-y.test)^2) 
pcr.mse = mean((pcr.pred-y.test3)^2) 
pls.mse = mean((pls.pred-y.test2)^2)
forests.mse = mean((yhat.rf-y.test4)^2)
gam.ns = mean((gam.pred-test_lm$y)^2)                                        
all.mse = c(lm.mse, forward.mse, backward.mse, ridge.mse, lasso.mse, pcr.mse, pls.mse,forests.mse,gam.ns)
names(all.mse) = c("lm", "forward", "backward" ,"ridge", "lasso", "pcr", "pls","Forests","GAM")
barplot(all.mse,ylim = c(1.45,1.65))

lm.mse
forward.mse
backward.mse
ridge.mse
lasso.mse
pcr.mse
pls.mse
forests.mse
gam.ns
```




FINAL MODEL;

```{r}
final_model = lm(y ~ month + lag1 + lag2 + aveLandTemp, data=data)
summary(final_model)
anova(final_model)

mean(final_model$residuals^2)       #final mse

res = resid(final_model)
plot(fitted(final_model), res)
abline(0,0)

qqnorm(res)
qqline(res)
```





Annual ave. temp by month;

```{r}
y1 <- data$y[10:nrow(data)]
year1 <- data$year[10:nrow(data)]
month1 <- data$month[10:nrow(data)]

data2 <- data.frame(y1,year1,month1)
data2$month1 <- as.factor(data2$month1)
data2$year1 <- as.numeric(as.character(data2$year1))
data2 = head(data2,-9)

l1 = nrow(data2)

l = l1/12

origional = data2$y1
p = 12
new = NULL
for (i in 1:l) {
  ydata = origional[((i-1)*p+1):(i*p)]
  yearlyAve = mean(ydata)
  new = rbind(new, yearlyAve)
}
ave_y1 = new

ts.plot(ave_y1)

yrs = rep(1850+1:length(ave_y1))

ave_y1 = noquote(cbind(yrs, ave_y1))

plot(ave_y1, type = "l", main = "Yearly Average California Temp.", xlab = "Year", ylab = "Temperature")

ave_y1

temp_1951_c = 14.14358
temp_2012_c = 15.35850

temp_2012_c-temp_1951_c

temp_1951_f = 57.458444
temp_2012_f = 59.6453

temp_2012_f-temp_1951_f
```




Visuals; 

```{r}
#-----------------------------------------------------------------------------------------------------------------------#

plot(data_final$year, data_final$y, xlim = c(1849, 2015))     #Plot Cali Average Temp ~ year  ? Uniform Data?
abline(0, 14.35696)

plot(factor(data_final$month), data_final$y)                  #box-Plot Cali Average Temp ~ month

plot(data_final$aveLandTemp, data_final$y)                    #Plot Cali Average Temp ~ average land temp  ? Not independent ?



#---------------------------------data_final Data-----------------------------------------#

hist(data$y, xlab = "Ave. California Temps", main = "Histogram of Ave. California Temps")           
hist(data_final$aveLandTemp, xlab = "Ave. California land Temps", main = "Histogram of Ave. California land Temps")             
hist(data_final$lag1)                                                         #Histogram LAG1
hist(data_final$lag2)                                                         #Histogram LAG2
hist(cali_final_data$LandAndOceanAverageTemperature, xlab = "Ave. California land and Ocean Temps", main = "Histogram of Ave. California land and ocean Temps")

plot(data_final)                                              #Scatterplot data for all variables 

summary(data_final)                                           #5 number summary for data

summary(cali_final_data)


#Plot Cali Average Temp ~ year
ggplot(data, aes(x = year, y = y)) + 
  geom_line() + 
  geom_smooth(method = "loess") +
  labs(title = " Cali Average Temp ~ year", y = "Cali Average Temp", x = "Year")


#Box-Plot Cali Average Temp ~ month
ggplot(data, aes(x = factor(month), y = y)) +
  geom_boxplot() +
  geom_smooth(method = "loess") +
  labs(title = "Cali Average Temp ~ month", y = "Cali Average Temp", x = "Month")


#Plot Cali Average Land Temp ~ year
ggplot(data, aes(x = year, y = aveLandTemp)) + 
  geom_line() + 
  geom_smooth(method = "loess") +
  labs(title = " Cali Average Land Temp ~ year", y = "Cali Average Land Temp", x = "Year")

#Plot Cali Average Land and Water Temp ~ year
#ggplot(cali_final_data, aes(x = year, y = LandAndOceanAverageTemperature)) + 
  #geom_line() + 
  #geom_smooth(method = "loess") +
  #labs(title = " Cali Average Land and Water Temp ~ year", y = "Cali Average Land and Water Temp", x = "Year")


#Box-Plot Cali Average Land Temp ~ month
ggplot(data, aes(x = factor(month), y = aveLandTemp)) + 
  geom_boxplot() +
  geom_smooth(method = "loess") +
  labs(title = " Cali Average Land Temp ~ month", y = "Cali Average Land Temp", x = "Month")


#Plot Cali Average Temp ~ year grouped by month
ggplot(data, aes(x = year, y = y, group = as.numeric(month))) + 
  geom_line(aes(colour = month)) +
  scale_color_gradient(low="blue", high="red") +
  labs(title = "Cali Average Temp ~ year grouped by month", y = "Average California Temp", x = "Year")


#Plot Cali Average Temp ~ month grouped by year
ggplot(data, aes(x = factor(month), y = y, group = year)) + 
  geom_line() +
  scale_x_discrete(labels = month.abb[1:12]) +
  labs(title = "Cali Average Temp ~ month grouped by year", y = "Average California Temp", x = "Months")


#Plot Cali Average Temp ~ year grouped by month
ggplot(data, aes(year, y, colour = as.numeric(month))) +
  geom_point(size = 0.5) +
  geom_smooth(method = "loess") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        axis.title.x = element_blank(),
        legend.position = "none") +
  scale_color_gradient(low="blue", high="red") +
  labs(title = "Cali Average Temp ~ year grouped by month", y = "Degrees Celsius") +
  facet_wrap(~month) +
  NULL

#Plot Cali Average Land Temp ~ year grouped by month
ggplot(data, aes(year, aveLandTemp, colour = as.numeric(month))) +
  geom_point(size = 0.5) +
  geom_smooth(method = "loess") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        axis.title.x = element_blank(),
        legend.position = "none") +
  scale_color_gradient(low="blue", high="red") +
  labs(title = "Cali Average Land Temp ~ year grouped by month", y = "Degrees Celsius") +
  facet_wrap(~month) +
  NULL

#Plot Cali Average Land and Water Temp ~ year grouped by month
ggplot(cali_final_data, aes(year, LandAndOceanAverageTemperature, colour = month)) +
  geom_point(size = 0.5) +
  geom_smooth(method = "loess") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        axis.title.x = element_blank(),
        legend.position = "none") +
  scale_color_gradient(low="blue", high="red") +
  labs(title = "Cali Average Land and Water Temp ~ year grouped by month", y = "Degrees Celsius") +
  facet_wrap(~month) +
  NULL


#---------------------------------cali_summer Data-----------------------------------------#

#Plot Cali Summer Average Temp ~ year
ggplot(cali_summer, aes(x = year, y = y, colour = month)) + 
  geom_line() + 
  geom_point() + 
  geom_smooth(method = "loess") +
  scale_color_gradient(low="blue", high="red") +
  labs(title = "Cali Summer Average Temp (June, July, Aug.) ~ year", y = "Cali Summer Average Temp", x = "Year")


#Box-Plot Cali Summer Average Temp ~ month
ggplot(cali_summer, aes(x = factor(month), y = y)) + 
  geom_boxplot() + 
  geom_smooth(method = "loess") +
  labs(title = "Cali Summer Average Temp (June, July, Aug.) ~ month", y = "Cali Summer Average Temp", x = "Month")


#Plot Cali Summer Average Land Temp ~ year
ggplot(cali_summer, aes(x = year, y = aveLandTemp)) + 
  geom_line() + 
  geom_smooth(method = "loess") +
  labs(title = " Cali Summer Average Land Temp ~ year", y = "Cali Summer Average Land Temp", x = "Year")


#Box-Plot Cali Summer Average Land Temp ~ month
ggplot(cali_summer, aes(x = factor(month), y = aveLandTemp)) + 
  geom_boxplot() +
  geom_smooth(method = "loess") +
  labs(title = " Cali Summer Average Land Temp ~ month", y = "Cali Summer Average Land Temp", x = "Month")


#Plot Cali Summer Average Temp ~ year grouped by month 
ggplot(cali_summer, aes(year, y, colour = month)) +
  geom_point(size = 0.5) +
  geom_smooth(method = "loess") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        axis.title.x = element_blank(),
        legend.position = "none") +
  scale_color_gradient(low="blue", high="red") +
  labs(title = "Summer Cali Average Temp ~ year grouped by month", y = "Degrees Celsius") +
  facet_wrap(~month) +
  NULL


#---------------------------------cali_winter Data-----------------------------------------#

#Plot Cali Winter Average Temp ~ year
ggplot(cali_winter, aes(x = year, y = y, colour = month)) + 
  geom_line() + 
  geom_point() + 
  geom_smooth(method = "loess") +
  scale_color_gradient(low="blue", high="red") +
  labs(title = "Cali Winter Average Temp (Dec, Jan, Feb.) ~ year", y = "Cali Winter Average Temp", x = "Year")


#Box-Plot Cali Winter Average Temp ~ month
ggplot(cali_winter, aes(x = factor(month), y = y)) + 
  geom_boxplot() + 
  geom_smooth(method = "loess") +
  labs(title = "Cali Winter Average Temp (Dec, Jan, Feb.) ~ month", y = "Cali Winter Average Temp", x = "Month")


#Plot Cali Winter Average Land Temp ~ year
ggplot(cali_winter, aes(x = year, y = aveLandTemp)) + 
  geom_line() + 
  geom_smooth(method = "loess") +
  labs(title = " Cali Winter Average Land Temp ~ year", y = "Cali Winter Average Land Temp", x = "Year")


#Box-Plot Cali Winter Average Land Temp ~ month
ggplot(cali_winter, aes(x = factor(month), y = aveLandTemp)) + 
  geom_boxplot() +
  geom_smooth(method = "loess") +
  labs(title = " Cali Winter Average Land Temp ~ month", y = "Cali Winter Average Land Temp", x = "Month")

#Plot Cali Winter Average Temp ~ year grouped by month 
ggplot(cali_winter, aes(year, y, colour = month)) +
  geom_point(size = 0.5) +
  geom_smooth(method = "loess") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        axis.title.x = element_blank(),
        legend.position = "none") +
  scale_color_gradient(low="blue", high="red") +
  labs(title = "Winter Cali Average Temp ~ year grouped by month", y = "Degrees Celsius") +
  facet_wrap(~month) +
  NULL


#---------------------------------Not Using-----------------------------------------#
#Plot Cali Average Temp ~ year
#ggplot(cali_final_data, aes(x = dt, y = AverageTemperature, group = 1)) + 
 # geom_line() + 
  #geom_smooth(method = "loess")

#Plot Cali Average Temp ~ all months of June
#ggplot(june, aes(x = year, y = y)) + 
  #geom_line() + 
  #geom_point() + 
  #geom_smooth(method = "loess")    

#Plot Cali Average Temp ~ all months of July
#ggplot(july, aes(x = year, y = y)) + 
  #geom_line() + 
  #geom_point() + 
  #geom_smooth(method = "loess")   

#Plot Cali Average Temp ~ all months of Aug.
#ggplot(aug, aes(x = year, y = y)) + 
  #geom_line() + 
  #geom_point() + 
  #geom_smooth(method = "loess")      
```




